---
author: Mojisoluwa (Jiso) Awe
title: "Operation Sticky Gate"
output: html_notebook
---

# What is Operation Sticky Gate?

Major League Baseball's 2021 season has been marked by record breaking statistics. Particularly the record low ERA's, extremely high spin rates, and a record breaking amount of no-hitters a few months into the season. Baseball is a game many consider boring from lack of action...
-avg spin rate prior to June 1st was around 2280
-mention the og steroid era
-data from start of season to June 1, and June 1 to July 17, from Baseball Savant.

# Before the Crackdown

For a long time MLB teams had what some would call a gentleman's agreement over utilizing foreign substances. Initially this started as a way for pitchers to better their grip on the ball when they pitched. Yet, now it's gotten to the point that the foreign substances are so sticky that instead of just helping the pitchers grip the ball, it's making the ball stick to their hand for longer, thus increasing the spin rate of the ball and making it harder to hit. As the spin of the ball increases the general location in which the ball is getting thrown is higher, adding to the difficulties hitters are facing. The spin on pitches like the 4-seam fastballs is a back spin, which opposes the downward force of gravity on the ball (think of the term 'rising fastball'). 

The MLB made a decision coming out against foreign substances by stating that starting on June 1st umpires will begin checking for pitchers using foreign substances and players caught using them will be removed from the game, placed on a 10 game suspension, and fined. Many people have argued that these statements won't deter pitchers from utilizing foreign substances. Yet many analysts have been saying spin rates going down and batting  averages going up. In this notebook we will be examining the affect of the MLB's new implementation and examining changes with individual players.

The data referenced has been provided by Baseball Savant and Baseball Reference.



## Velocity vs Spin Rate

It's important to note that generally as increases velocity so does spin rate. In this section we will be comparing pitchers spin rates vs velocities before and after June 1st.

```{r}
library(ggplot2) ## loading packages
library(ggExtra)
library(dplyr)
library(tidyverse)
```



```{r}
b4 <-
  read.csv('b4June1.csv')
b4
```



```{r}
b4gr <-
  b4 %>%
  ggplot(aes(x = velocity, y = spin_rate, color = total_pitches)) +
  geom_point(stat = 'identity') +
  xlab('Velocity (mph)') +
  ylab('Spin Rate (rpm)') +
  ggtitle('Spin Rate VS Velocity Prior to June 1st') +
  labs(color = 'Total Pitches')

p1 <-
  ggMarginal(b4gr, type = 'boxplot')
p1
```

 * insert analysis *
 
 
 
 
```{r}
aft <-
  read.csv('aftJune1st.csv')
aft
```


```{r}
aftgr <-
  aft %>%
  ggplot(aes(x = velocity, y = spin_rate, color = total_pitches)) +
  geom_point(stat = 'identity') +
  xlab('Velocity (mph)') +
  ylab('Spin Rate (rpm)') +
  ggtitle('Spin Rate VS Velocity After June 1st') +
  labs(color = 'Total Pitches')

p2 <-
  ggMarginal(aftgr, type = 'boxplot')
p2
p1
```


Based on these general scatterplots there doesn't seem to be a large change between between spin rate and velocity. The boxplots on the side show that...
-most of the outliers are darker shaded dots showing that those pitchers generally haven't thrown as many pitches.
-Now let's change the range in the number of pitches to above 325....


```{r}
b4abv <-
  b4%>%
  filter(total_pitches > 249)
b4abv
```

```{r}
aftabv <-
  aft%>%
  filter(total_pitches > 249)
aftabv
```

```{r}
babvgr <-
  b4abv %>%
  ggplot(aes(x = velocity, y = spin_rate, color = total_pitches)) +
  geom_point(stat = 'identity')
babvgr
```


```{r}
afabvgr <-
  aftabv %>%
  ggplot(aes(x = velocity, y = spin_rate, color = total_pitches)) +
  geom_point(stat = 'identity')
afabvgr
```



## Individual Pitcher Changes

#### Trevor Bauer

It's hard to have any conversation about foreign substances in baseball without mentioning Trevor Bauer. From being someone who initially 

#### Jacob deGrom

#### Gerrit Cole

#### Julio Urias

#### Garrett Richards
```{r}

```


