---
author: Mojisoluwa (Jiso) Awe
title: "Operation Sticky Gate"
output: html_notebook
---

# What is Operation Sticky Gate?

Major League Baseball's 2021 season has been marked by record breaking statistics. Particularly the record low ERA's, extremely high spin rates, and a recording breaking amount of no-hitters a few months into the season. Many consider baseball boring due to a 'lack of action', well this season that's proving true. 
-avg spin rate prior to June 1st was around 2280
-mention the og steroid era
-data from start of season to June 1, and June 1 to July 17, from Baseball Savant.
-sinkers and curveballs

# Before the Crackdown

For a long time MLB teams had what some would call a gentleman's agreement over utilizing foreign substances. Initially this started as a way for pitchers to better their grip on the ball when they pitched. Now, it's gotten to the point that the foreign substances are so sticky that rather than just helping pitchers grip the ball, it's making the ball stick to their hand for longer, thus increasing the spin rate of the ball and making it harder to hit. As the spin of the ball increases the general location in which the ball is getting thrown is higher, adding to the difficulties hitters are already facing. The spin on pitches like the 4-seam fastballs is a back spin, which opposes the downward force of gravity on the ball (think of the term 'rising fastball'). 

The MLB made a decision coming out against foreign substances by stating that starting on June 1st umpires will begin checking for pitchers using foreign substances and on June 16th announced players caught using them will be removed from the game, placed on a 10 game suspension, and fined. Many people have argued that these statements won't deter pitchers from utilizing foreign substances. Yet many analysts have been saying spin rates going down and batting  averages going up. In this notebook we will be examining the affect of the MLB's new implementation and examining changes with individual players.

All the data included has been provided by Baseball Savant found through utilizing the search feature and Baseball Reference (linked when referenced).



## Velocity vs Spin Rate

It's important to note that generally as increases velocity so does spin rate. In this section we will be comparing pitchers spin rates vs velocities before and after June 1st.


```{r}
library(ggplot2) ## loading packages
library(ggExtra)
library(dplyr)
library(tidyverse)
library(lubridate)
```



```{r}
b4 <-
  read.csv('b4June1.csv')
b4
```


 
```{r}
aft <-
  read.csv('aftJune1st.csv')
aft
```


```{r}
b4gr <-
  b4 %>%
  ggplot(aes(x = velocity, y = spin_rate, color = total_pitches)) +
  geom_point(stat = 'identity') +
  xlab('Velocity (mph)') +
  ylab('Spin Rate (rpm)') +
  ggtitle('Spin Rate VS Velocity Prior to June 1st') +
  labs(color = 'Total Pitches')

##using ggExtra package to add histograms to show the distribution of data within the scatterplot
p1 <-
  ggMarginal(b4gr, type = 'histogram')
p1


aftgr <-
  aft %>%
  ggplot(aes(x = velocity, y = spin_rate, color = total_pitches)) +
  geom_point(stat = 'identity') +
  xlab('Velocity (mph)') +
  ylab('Spin Rate (rpm)') +
  ggtitle('Spin Rate VS Velocity After June 1st') +
  labs(color = 'Total Pitches')

p2 <-  
  ggMarginal(aftgr, type = 'histogram')
p2
```



The graph of the data before June 1st shows that generally pitchers average spin rates are concentrated between 2200 rpm and 2600 rpm and the velocities range from the upper 80s to mid 90s. 

The graph of the data following June 1st, displays a higher conglomeration of pitchers with average pitches between 2000 and 2500 rpm and less outliers with spin rates above 2750 rpm. With velocities concentrated between the upper 80's and lower 90's and spin rates between 2000 rpm and 2500 rpm.

Upon further inspection of the scatterplot it appears that most of the outliers are darker shaded dots representing pitcher who've thrown between 1 and 250 pitches. Because this data is based on averages and smaller sample sizes are generally less representative of the individual. This prompted me to see if there was a difference when if I increased the minimum number of pitches to at least 250.


```{r}
b4abv <-
  b4%>%
  filter(total_pitches > 249)
b4abv
```

```{r}
aftabv <-
  aft%>%
  filter(total_pitches > 249)
aftabv
```




```{r}
babvgr <-
  b4abv %>%
  ggplot(aes(x = velocity, y = spin_rate, color = total_pitches)) +
  geom_point(stat = 'identity') +
  xlab('Velocity (mph)') +
  ylab('Spin Rate (rpm)') +
  labs(color = 'Total Pitches',
       title = 'Spin Rate vs Velocity Before June 1st',
       caption = 'For pitchers with over 250 pitches') +
  scale_color_viridis_c()
g1 <-
  ggMarginal(babvgr, type = 'histogram')
g1


afabvgr <-
  aftabv %>%
  ggplot(aes(x = velocity, y = spin_rate, color = total_pitches)) +
  geom_point(stat = 'identity') +
  xlab('Velocity (mph)') +
  ylab('Spin Rate (rpm)') +
  labs(color = 'Total Pitches',
       title = 'Spin Rate vs Velocity After June 1st',
       caption = 'For pitchers with over 250 pitches') +
  scale_color_viridis_c()

k1 <- 
  ggMarginal(afabvgr, type = 'histogram')
k1
```

Here we can see a more notable difference between average spin rates and velocities. Prior to June 1st we can see that average velocities were generally between the upper 80s and mid 90s, with spin rates ranging between 2100 and 2600 rpm. The extremes included 5 pitchers with spin rates above 2750 rpm and 20 pitchers above 95 mph. 
In contrast we can see that after June 1st, velocities are concentrated between the mid 80s to lower 90's and spin rates between 2000 and 2500 rpm. The extremes here now only include 1 pitcher with an average spin rate above 2750 and 3 pitchers above 95 mph. Additionally the frequency distribution of the velocities changed from left skew to a normal distribution. Whereas the frequency distribution of the spin rates went from bi-modal to a normal distribution following June 1st.


# Individual Pitcher Changes

In this section we will be examining a change in spin rate, velocity, and overall performance (if any) between specific pitchers: Trevor Bauer, Gerrit Cole, Jacob deGrom, and Garrett Richards.

## Trevor Bauer

It's hard to have any conversation about foreign substances in baseball without mentioning Trevor Bauer. From being someone who initially spoken against foreign substances and once even stating that they could be more powerful than steroids, to now being accused of having the most effective sticky substance combination in the league, Bauer's name has been brought up a lot. Considering recent events regarding Bauer's arrest and allegation, the data available is more limited than with other players.

```{r}
bauer <- ## downloaded from Baseball Savant
  read.csv('bauer advanced stats.csv')
bauer
```


```{r}
num <-  ## finding counts of each pitch type
  bauer %>%
  group_by(pitch_name) %>%
  summarise(count = n())
num
```


From this data we can see that during the month of June, Bauer's most utilized pitches were the fastball, cutter, and slider.


#### June 28th

```{r}
tbju28 <-
  bauer %>%
  filter(game_date == '2021-06-28')
tbju28
```

```{r}
tb2avgs <- ## finding game averages from the June 28th game
  tbju28 %>%
  select(pitch_name, release_speed, release_spin_rate) %>%
  group_by(pitch_name) %>%
  summarise(avgspeed = mean(release_speed, na.rm = TRUE), avgspin = mean(release_spin_rate, na.rm = TRUE), count = n()) %>%
  arrange(pitch_name)
tb2avgs
```



From this we can see that Bauer's fastest average pitches are his sinker, fastball, and changeup. His pitches with the highest spin rates are the knuckle curve, slider, cutter, and fastball, all with spin rates all well above the average spin rate of 2280 rpm.

Bauer's most utilized pitches this game:

* Cutter (medium velocity & high spin rate)

* Slider (low velocity & high spin rate)

* Fastball (high velocity & spin rate)


note: Bauer's ERA during this game was 2.59 [Baseball Reference](https://www.baseball-reference.com/boxes/LAN/LAN202106280.shtml)

#### June 6th

```{r}
tbju6 <-
  bauer %>%
  filter(game_date == '2021-06-06')
tbju6
```


```{r}
tb6avgs <-
  tbju6%>%
  select(pitch_name, release_speed, release_spin_rate) %>%
  group_by(pitch_name) %>%
  summarise(avgspeed = mean(release_speed, na.rm = TRUE), avgspin = mean(release_spin_rate, na.rm = TRUE), count = n()) %>%
  arrange(pitch_name)
tb6avgs
```

Here we can see that from Bauer's June 6th game, the pitches with the highest speed were the fastball, sinker, and changeup. And his pitches with the highest spin rates were the knuckle curveball, slider, cutter, sinker, and fastball, all significantly above league average of 2280 rpm.


Bauer's most utilized pitches during this game:

* Fastball (high velocity & high spin rate)
* Cutter (medium velocity & high spin rate)
* Slider (low velocity & high spin rate)


note: Bauer's ERA during from this game was 2.40 [Baseball Reference](https://www.baseball-reference.com/boxes/ATL/ATL202106060.shtml)

```{r}
tb6gr <-    ## depicting Bauer's spin rates by pitch from June 6th
  tbju6%>%
  ggplot(aes(x = release_speed , y = release_spin_rate, color = pitch_name)) +
  geom_point(stat = 'identity')+
  xlab('Release Speed (mph)') +
  ylab('Release Spin Rate (rpm)') +
  labs(title = "Trevor Bauer's Spin Rate by Pitch Type",
       color = 'Pitch Name',
       caption = 'From game pitched on June 6th')
tb6gr

tb28gr <-  ## depicting Bauer's spin rates by pitch from June 28th
  tbju28%>%
  ggplot(aes(x = release_speed , y = release_spin_rate, color = pitch_name)) +
  geom_point(stat = 'identity')+
  xlab('Release Speed (mph)') +
  ylab('Release Spin Rate (rpm)') +
  labs(title = "Trevor Bauer's Spin Rate by Pitch Type",
       color = 'Pitch Name',
       caption = 'From his last game on June 28th')
tb28gr
```


Here we can see the drastic drop in Bauer's spin rate between the two games only 22 days apart. During Bauer's June 6th game, before the League started implementing the 10-day suspension rule, it can be seen that he has several pitches being thrown at above 3000 rpm and with the exception of his changeup's all of his pitches are rotating above 2500 rpm. Yet, by June 28th, Bauer didn't have a single pitch with a spin rate above 3000 rpm. In addition to this, Bauer's [ERA](https://github.com/jisoawe/MLB-Data-Visualizations/blob/main/ERA%20Explained.Rmd) during the two games increased from 2.40 to 2.59, meaning he was allowing significantly more runs and hits per inning after implementation. When seeing such large drops in spin rates, it can be speculated that the pitcher was utilizing sticky substances.

#### Bauer's Sticky Verdict: GUILTY



## Gerrit Cole

Yankee's pitcher Gerrit Cole has been leading the league in strikeouts for the entire season. And after [this](https://twitter.com/mroberson22/status/1402377614086467586?s=20) infamous post game interview of Gerrit Cole, I can't not look at his stats and see if the stuttering was an admission of guilt or if he was just tired. 


```{r}
gcoledt <-
  read.csv('Gerrit Cole advanced stats.csv')
gcoledt
```

```{r}
gccounts <-
  gcoledt %>%
  group_by(pitch_name) %>%
  summarise(count = n())
gccounts
```


From this we can see that Cole's most utilized pitch by far is his fastball, followed by his slider, knuckle curve, and changeup. Whereas his sinker is rarely used, only making an appearance 8 times in almost 800 pitches tracked.



```{r}
gcju3 <-  ## data from Cole's June 3rd game
  gcoledt %>%
  filter(game_date == '2021-06-03')
gcju3
```


```{r}
ju3avg <-
  gcju3%>%
  select(pitch_name, release_spin_rate, release_speed) %>%
  group_by(pitch_name) %>%
  summarise(avgspeed = mean(release_speed, na.rm = TRUE), avgspin = mean(release_spin_rate, na.rm = TRUE), count = n()) %>%
  arrange(pitch_name)
ju3avg
```


We can see that from Cole's June 3rd game, the pitches with the highest speed were the fastball, sinker, changeup, and slider. And his pitches with the highest spin rates were the knuckle curve, slider, fastball, and sinker, with all being at least 100 rpm above the league average of 2280 rpm.


Cole's most utilized pitches during this game:

* Fastball (high velocity & high spin rate)

* Knuckle Curve (medium velocity & high spin rate)

* Changeup (high velocity & low spin rate)


note: Cole's ERA during from this game was 2.26 [Baseball Reference](https://www.baseball-reference.com/boxes/NYA/NYA202106030.shtml)


```{r}
gcjl17 <-
  gcoledt %>%
  filter(game_date == '2021-07-17')
gcjl17
```



```{r}
jl17avg <-
  gcjl17 %>%
  group_by(pitch_name) %>%
  select(pitch_name, release_spin_rate, release_speed) %>%
  summarise(avgspin = mean(release_spin_rate, na.rm = TRUE), avgspeed = mean(release_speed, na.rm = TRUE), count = n()) %>%
  arrange(pitch_name)
jl17avg
```


Here we can see that from Cole's July 17th game, the pitches with the highest speed were the fastball, changeup, and slider. And his pitches with the highest spin rates were the knuckle curve, slider, and fastball, all still well above the league average of 2280 rpm.


Cole's most utilized pitches during this game:

* Fastball (high velocity & medium spin rate)
* Knuckle Curve (medium velocity & high spin rate)
* Slider (medium velocity & high spin rate)

note: Cole's ERA during from this game was 2.63 [Baseball Reference](https://www.baseball-reference.com/boxes/NYA/NYA202107170.shtml)


```{r}
gcju3gr <-
  gcju3 %>%
  ggplot(aes(x = release_speed, y = release_spin_rate, color = pitch_name)) +
  geom_point(stat = 'identity') +
  xlab('Release Speed (mph)') +
  ylab('Release Spin Rate (rpm)') +
  labs(title = "Gerrit Cole's Spin Rate by Pitch",
       caption = 'From the game pitched on June 3rd',
       color = 'Pitch Type')+
  scale_color_manual(values = c('4-Seam Fastball' = 'chocolate1',
                                'Changeup' = 'darkgoldenrod1',
                                'Knuckle Curve' = 'darkseagreen3',
                                'Sinker' = 'darkorchid1',
                                'Slider' = 'deepskyblue'))
gcju3gr


gcjl17gr <-
  gcjl17 %>%
  ggplot(aes(x = release_speed, y = release_spin_rate, color = pitch_name)) +
  geom_point(stat = 'identity') +
  xlab('Release Speed (mph)') +
  ylab('Release Spin Rate (rpm)') +
  labs(title = "Gerrit Cole's Spin Rate by Pitch Type",
       caption = 'From the game Cole pitched on July 17th',
       color = 'Pitch Type') +
  scale_color_manual(values = c('4-Seam Fastball' = 'chocolate1',
                                'Changeup' = 'darkgoldenrod1',
                                'Knuckle Curve' = 'darkseagreen3',
                                'Slider' = 'deepskyblue'))
gcjl17gr
```

When comparing the two graphs the change in Cole's spin rates isn't as drastic as Bauer's. His changeup, fastball, and slider remained relatively similar to what they were during his starts from his games prior to implementation. Additionally, the spin on his curveball increased, nearly reaching 3000 rpm during at bat number 40. His average release velocities also stayed relatively similar. Additionally, Cole's ERA increased by 0.4 between during this these two games, leading me to think that with his pitches aren't rising up as much, therefore making them more hittable.


#### Cole's Sticky Verdict: uh well... (probably) GUILTY

## Jacob deGrom

Jacob deGrom is arguably one of the best pitchers in all of baseball. With two Cy Young awards, a season ERA of 1.08, a season WAR of 5.0 (as of July 29th, 2021), and considering he was a shortstop in college, there is no doubt that he is a gifted player. Following all of his stellar performances this season, I was wondering if he had any change in pitching output.


```{r}
dgstats <-
  read.csv('deGrom Advanced Stats.csv')
dgstats
```


```{r}
ap17 <-  ##filtering data from deGrom's first game
  dgstats %>%
  filter(game_date == '2021-04-17')
ap17
```

```{r}
ap17avg <-
  ap17 %>%
  group_by(pitch_name) %>%
  select(pitch_name, release_spin_rate, release_speed) %>%
  summarise(avgspin = mean(release_spin_rate, na.rm = TRUE), avgspeed = mean(release_speed, na.rm = TRUE), count = n()) %>%
  arrange(pitch_name)
ap17avg
```

From this data we can see that during this game deGrom only used three pitches, his fastball, changeup, and slider. All of these pitches had high speeds, each averaging above 90 mph. The fastball was used the most with that accounting for almost 60% of his pitches, whereas the changeup accounts for only 9% of his pitches from this game.

deGrom's utilized pitches during this game:

* Fastball (high velocity & medium spin rate)
* Slider (high velocity & high spin rate)
* Changeup (high velocity & low spin rate)


note: deGrom's ERA during from this game was 0.45 [Baseball Reference](https://www.baseball-reference.com/boxes/COL/COL202104171.shtml)


```{r}
ap17gr <-
  ap17 %>%
  ggplot(aes(x = release_speed, y = release_spin_rate, color = pitch_name)) +
  geom_point(stat = 'identity') +
  xlab('Release Speed (mph)') +
  ylab('Release Spin Rate (rpm)') +
  labs(title = "Jacob deGrom's Spin Rate by Pitch",
       caption = 'From the game pitched on April 17th',
       color = 'Pitch Type')
ap17gr
```







```{r}
l8r <-
  dgstats %>%
  group_by(pitch_name) %>%
  summarise(count = n())
l8r
```

We can see that deGrom's most utilized this season have been the fastball, slider, and changeup, with his curveball only making 4 appearances.


```{r}
stpg <-
  dgstats %>%
  select(game_date, release_spin_rate, pitch_name) %>%
  group_by(game_date, pitch_name) %>%
  summarise(avgspin = mean(release_spin_rate), pitch_name, game_date)%>%
  summarise(avgspin, pitch_name, game_date) %>%
  mutate(game_date = date(game_date)) %>%
  filter(pitch_name != ('Curveball')) %>%
  filter(pitch_name != ('Changeup'))
np <- stpg[!duplicated(stpg),]
np

slst <-
  dgstats %>%
  select(game_date, release_spin_rate, pitch_name) %>%
  group_by(game_date, pitch_name) %>%
  summarise(avgspin = mean(release_spin_rate)) %>%
  mutate(game_date = date(game_date))%>%
  filter(pitch_name == 'Changeup')
lp <- slst[!duplicated(slst),]
lp
```


```{r}
idkyet <-
  np %>%
  ggplot(aes(fill = fct_rev(pitch_name), y = avgspin, x = game_date)) +
  geom_bar(position = 'stack', stat = 'identity') +
  xlab('Game Date') +
  ylab('Average Spin Rate by Pitch') +
  labs(fill = 'Pitch Type',
       caption = "From the start of the season to his last game on July 7th",
       title = "deGrom's Average Spin Rate by Game")
idkyet

cug <-
  lp %>%
  ggplot(aes(x = game_date, y = avgspin )) +
  geom_bar(stat = 'identity', fill = 'palegreen3') +
  xlab('Game') +
  ylab('Average Spin Rate') +
  labs(title = "deGrom's Average Changeup Spin Rate by Game",
       caption = "From the beginning of the season to July 7th")
cug
```

From this data we can see that deGrom's average spin rates from the start of the season to July 7th, that there is no large change in spin rate for any of his pitches. There seems to be more of a change when examining deGrom's changeups. They're all generally between 1500 and 1750 rpm, yet during his July 7th appearance his average jumps up to 1887.25 rpm. Due to deGrom utilizing his changeup only 104 times during this time period, I made a separate graph displaying those averages. This lack of change, leads me to believe that deGrom wasn't excessively using foreign substances.

####  deGrom's Sticky Verdict: NOT Guilty


## Garrett Richards

Garrett Richards', Boston's newest pitcher has had a very [word] season so far...

```{r}
grichdt <-  ## Garrett Richards data from Baseball Savant
  read.csv('Garrett Richards advanced stats.csv')
grichdt
```


```{r}
grcounts <-
  grichdt %>%
  group_by(pitch_name) %>%
  summarise(count = n())
grcounts
```

Here we can see that Richards' most utilized pitch is his fastball with over 900 appearances within between May 8th and July 9th, followed by his slider and curveball. Whereas he rarely threw changeup and sinkers, only being used 33 and 6 times respectively.



```{r}
gravs <-
  grichdt %>%
  select(game_date, release_spin_rate, pitch_name) %>%
  group_by(game_date, pitch_name) %>%
  summarise(avgspin = mean(release_spin_rate, na.rm = TRUE)) %>%
  mutate(game_date = date(game_date))
gravs

grfbs <-
  grichdt %>%
  filter(pitch_name == '4-Seam Fastball' | pitch_name == 'Slider') %>%
  select(game_date, release_spin_rate, pitch_name) %>%
  group_by(game_date, pitch_name) %>%
  summarise(avgspin = mean(release_spin_rate, na.rm = TRUE)) %>%
  mutate(game_date = date(game_date))
grfbs

grcb <-
  grichdt %>%
  filter(pitch_name == 'Curveball') %>%
  select(game_date, release_spin_rate, pitch_name) %>%
  group_by(game_date, pitch_name) %>%
  summarise(avgspin = mean(release_spin_rate, na.rm = TRUE)) %>%
  mutate(game_date = date(game_date))
grcb
```


```{r}
gravggr <-
  gravs %>%
  ggplot(aes(x = game_date, y = avgspin, fill = pitch_name)) +
  geom_bar(stat = 'identity', position = 'stack')
gravggr

grfbsgr <-
  grfbs%>%
  ggplot(aes(x = game_date, y = avgspin, fill = pitch_name)) +
  geom_bar(stat = 'identity', position = 'stack') +
  xlab('Game Date') +
  ylab('Average Spin Rate (RPM)') +
  labs(title = "Garrett Richard's Average Fastball and Slider Spin Rates",
       caption = 'Data')
grfbsgr

grcbgr <-
  grcb%>%
  ggplot(aes(x = game_date, y = avgspin)) +
  geom_bar(stat = 'identity', fill = 'palegreen3') +
  ylab('Average Spin Rate (rpm)') +
  xlab('Game Date') +
  labs(title = "Garrett Richard's Average Curveball Spin Rate",
       caption = 'From the start of the season to July 7th')
grcbgr
```



```{r}
grjul9 <-  ## filtering the data from Garrett Richard's most recent game
  grichdt %>%
  filter(game_date == '2021-07-09')
grjul9
```


```{r}
jul9avgs <-  ##averages from July 9th game
  grjul9%>%
  select(pitch_name, release_spin_rate, release_speed) %>%
  group_by(pitch_name) %>%
  summarise(avgspeed = mean(release_speed, na.rm = TRUE), avgspin = mean(release_spin_rate, na.rm = TRUE), count = n())
jul9avgs
```

Here we can see that from Richards' July 9th game, the pitches with the highest speeds were the fastball and changeup. And his pitches with the highest spin rates were the curveball, slider, and fastball, all above the league average of 2280 rpm.


Richards' most utilized pitches during this game:

* Fastball (high velocity & medium spin rate)

* Curveball (low velocity & high spin rate)

* Slider (medium velocity & high spin rate)

note: Richards' ERA during from this game was 4.91 [Baseball Reference](https://www.baseball-reference.com/boxes/BOS/BOS202107090.shtml)

```{r}
grmay8 <-  ## filtering the data from Richard's May 8th game
  grichdt %>%
  filter(game_date == '2021-05-08')
grmay8
```


```{r}
may8avgs <-
  grmay8%>%
  select(release_spin_rate, release_speed, pitch_name) %>%
  group_by(pitch_name) %>%
  summarise(avgspeed = mean(release_speed, na.rm = TRUE), avgspin = mean(release_spin_rate, na.rm = TRUE), count = n())
may8avgs
```

During Richards' May 8th game, it can seen he only utilized three different types of pitches. The two with the highest speeds were the fastball and slider. The two pitch types with the highest spin rates were the curveball and slider. Note all of his average spin rates from this game which were significantly above the league average of 2280 rpm.


Richards' most utilized pitches during this game:

* Fastball (high velocity & high spin rate)

* Slider (medium velocity & high spin rate)

* Curveball (low velocity & high spin rate)

note: Richards' ERA during from this game was 4.54 [Baseball Reference](https://www.baseball-reference.com/boxes/BAL/BAL202105080.shtml)

```{r}
my8gr <-
  grmay8%>%
  ggplot(aes(x = release_speed, y = release_spin_rate, color = pitch_name))+
  geom_point() +
  xlab('Release Speed (mph)')+
  ylab('Spin Rate (rpm)') +
  labs(title = 'Gerritt Richards Spin Rate by Pitch',
       color = 'Pitch Type',
       caption = 'From game pitched on May 8th') +
  scale_color_manual(values = c('4-Seam Fastball' = 'chocolate1',
                                'Curveball' = 'cyan3',
                                'Slider' = 'darkorchid2'))
my8gr

grjul9gr <-
  grjul9 %>%
  ggplot(aes(x = release_speed, y = release_spin_rate, color = pitch_name)) +
  geom_point() +
  xlab('Release Speed (mph)') +
  ylab('Release Spin Rate') +
  labs(title = 'Gerritt Richards Spin Rate by Pitch',
       caption = 'For game pitched on July 9th',
       color = 'Pitch Type') +
  scale_color_manual(values = c('4-Seam Fastball' = 'chocolate1',
                                'Changeup' = 'chartreuse3',
                                'Curveball' = 'cyan3',
                                'Slider' = 'darkorchid2'))
grjul9gr
```


When examining Richards' pitch velocities and spin rates between these two games there is a clear difference in performance. Generally, his pitches are more concentrated in the July game, showing that there's more consistency between the pitches he was throwing in July than in May. Richards' went from having more than half of his curveballs above 3300 rpm in May to not even having a single curveball reach 3000 rpm. When comparing the averages there's been a drop of 410 rpm in this pitch and a 4.25 mph drop in velocity. When looking at his sliders, the spin rates are concentrated around 2900 rpm, which then dropped to around 2600 rpm. The velocity change with this pitch wasn't as drastic shifting with an average of about 1.9 mph, which wouldn't be very suspicious if not for the other changes. Additionally, when looking at his fastballs (his most utilized pitch) there isn't as significant of a change. The velocities are concentrated in the same area and only had a change in 0.45 mph. The average spin rates for the fastball only dropped by 175 rpm, which by itself isn't particularly suspicious. Yet when looking at all of the changes in Richards' pitches and his game specific ERA increasing by 0.63 there is more than ample evidence that Richards' was in fact utilizing sticky substances.



####  Richards' Sticky Verdict: GUILTY






